The Talking Heads experiment

### 2015-04-29 chap3/chap3.tex
part 1 of martin haspelmath's comments

\is{conceptual spaces}The sensory channels used in the experiments have
deliberately been kept to a minimum so that we can follow 
in detail the ontological and lexical dynamics, which is a 
non-trivial matter as I will demonstrate in Chapters 6 and 7.
Obviously to get a richer lexicon, more channels need to be made 
available to the agents. These channels can often be 
grouped to form a {\itshape conceptual space}.\footnote{See \cite{Gardenfors:1999}. 
More cognitive oriented spaces, further removed 
from perception, are discussed in \cite{Fauconnier:1994}.} One of the 
best known examples is the colour space formed by the
yellow-blue, red-green, black-white, saturation, and brightness channels.

@@

A first frame of reference is based on
the absolute minima and maxima of the values 
on a sensory channel. This can be used to do 
{\itshape sensor-oriented scaling}. For example, 
the image map captured by the camera
contains 320 × 240 pixels, which means 
that the horizontal position (\textsc{hpos}) has a value between
0 and 320 and the vertical position (\textsc{vpos}) has a value 
between 0 and 240. Both values can be scaled to fall between 
0.0 and 1.0 so that they become uniform with respect to 
other sensory channels. Given a value $x$ and a 
$min$ and $max$ value, then the scaled value is 
$x' = {(x - min)}/{(max - min)}$. 
Thus the sensed value of \textsc{hpos}=200 becomes
\textsc{hpos}=0.62 after scaling .

In the experiments discussed later, 
sensor-scaling is always performed so that all sensory data
is between 0.0 and 1.0, allowing values on different channels 
to be compared with 
each other. In addition, context-oriented scaling is performed.
{\itshape Context-oriented scaling} uses the minima and maxima of the 
values that effectively appear in the context.
For example, if the values for the \textsc{height} of segments 
observed in the scene is between 500 and 700, 
then these can be the minimum and maximum of the 
\textsc{height} scale, so that 500 maps onto 0.0 and 700 to 1.0. 
Context-oriented scaling acts like a lens
magnifying the perceived values so that distinctions
stand out more clearly. This scaling could 
be made more sophisticated by introducing an additional scaling factor so that 
differences are amplified but not necessarily blown up to their extremes.

@@

A final form of scaling uses the typical values of the object being 
perceived. I will call this {\itshape object-oriented scaling}. 
For example, a small elephant is always
very large next to a large mouse. We clearly
have expectations about the typical size of an elephant at
a certain distance and use it to scale
perceptual data prior to size 
categorisation. Context-oriented and object-oriented scaling
are examples how constraints must flow 
from non-visual cognitive processes to visual processing. 
These non-modular interactions are a strong
indication that cognitive subsystems must be highly 
interconnected.

@@

\begin{figure}[hbtp]
\begin{center}
\includegraphics[width=0.8\columnwidth]{chap7/figs/plate-10.pdf}
\end{center}
\caption{Three examples of segmented images. The 
topic is indicated by a dashed bounding box in the 
image of the speaker. Segments which are too small 
are ignored. The topics have all been conceptualised
as being to the right and so the same word 
\smplenquote{gofubo} has been used to refer to them. }
\label{f:plate-10}
\end{figure}

>>>>

\is{conceptual spaces}The sensory channels used in the experiments have
deliberately been kept to a minimum so that we can follow 
in detail the ontological and lexical dynamics, which is a 
non-trivial matter as I will demonstrate in Chapters 6 and 7.
Obviously to get a richer lexicon, more channels need to be made 
available to the agents. These channels can often be 
grouped to form a {\scshape conceptual space}.\footnote{See \cite{Gardenfors:1999}. 
More cognitive oriented spaces, further removed 
from perception, are discussed in \cite{Fauconnier:1994}.} One of the 
best known examples is the colour space formed by the
yellow-blue, red-green, black-white, saturation, and brightness channels.

@@

A first frame of reference is based on
the absolute minima and maxima of the values 
on a sensory channel. This can be used to do 
{\scshape sensor-oriented scaling}. For example, 
the image map captured by the camera
contains 320 × 240 pixels, which means 
that the horizontal position (\textsc{hpos}) has a value between
0 and 320 and the vertical position (\textsc{vpos}) has a value 
between 0 and 240. Both values can be scaled to fall between 
0.0 and 1.0 so that they become uniform with respect to 
other sensory channels. Given a value $x$ and a 
$min$ and $max$ value, then the scaled value is 
$x' = {(x - min)}/{(max - min)}$. 
Thus the sensed value of \textsc{hpos}=200 becomes
\textsc{hpos}=0.62 after scaling .

In the experiments discussed later, 
sensor-scaling is always performed so that all sensory data
is between 0.0 and 1.0, allowing values on different channels 
to be compared with 
each other. In addition, context-oriented scaling is performed.
{\scshape Context-oriented scaling} uses the minima and maxima of the 
values that effectively appear in the context.
For example, if the values for the \textsc{height} of segments 
observed in the scene is between 500 and 700, 
then these can be the minimum and maximum of the 
\textsc{height} scale, so that 500 maps onto 0.0 and 700 to 1.0. 
Context-oriented scaling acts like a lens
magnifying the perceived values so that distinctions
stand out more clearly. This scaling could 
be made more sophisticated by introducing an additional scaling factor so that 
differences are amplified but not necessarily blown up to their extremes.

@@

A final form of scaling uses the typical values of the object being 
perceived. I will call this {\scshape object-oriented scaling}. 
For example, a small elephant is always
very large next to a large mouse. We clearly
have expectations about the typical size of an elephant at
a certain distance and use it to scale
perceptual data prior to size 
categorisation. Context-oriented and object-oriented scaling
are examples how constraints must flow 
from non-visual cognitive processes to visual processing. 
These non-modular interactions are a strong
indication that cognitive subsystems must be highly 
interconnected.

@@

\begin{figure}[hbtp]
\begin{center}
\includegraphics[width=0.8\columnwidth]{chap7/figs/plate-10.pdf}
\end{center}
\caption{Three examples of segmented images. The 
topic is indicated by a dashed bounding box in the 
image of the speaker. Segments which are too small 
are ignored. The topics have all been conceptualised
as being to the right and so the same word 
\textit{gofubo} has been used to refer to them. }
\label{f:plate-10}
\end{figure}

### 2015-04-29 chap4/chap4.tex
part 1 of martin haspelmath's comments

The previous chapter laid the perceptual groundwork 
for investigating conceptualisation and language 
communication. It proposed a set of mechanisms for segmenting a 
scene and extracting characteristics about each 
segment. The next task of the agent is to use these results
from the perceptual layer to categorise 
and hence conceptualise the
scene. For the time being, I will 
ignore the full complexity of natural
language meaning and focus only on the most 
simple type of conceptualisation one can
imagine, namely categories, which logicians refer 
to as unary predicates, as well as conjunctive 
combinations of categories. Words like \smplenquote{blue}, \smplenquote{light} or
\smplenquote{square} name such categories. We not only need a way 
that the Talking Heads can use such categories to conceptualise
a scene based on visual perception, but we also need 
mechanisms that can explain how such categories might 
develop or be acquired by each agent autonomously and 
without supervised training. This is clearly an enormous
challenge and no known universally accepted solution 
exists to this problem.

@@

Then there is a third paradox, first suggested
by Jean-Jacques Rousseau. In order to communicate,
the speaker and the hearer need to share the building
blocks of the conceptualisations underlying their communication. 
If I say \dblenquote{the wine glass on the table}, I expect 
the hearer to be able to recognise what objects are 
tables, when something is on the table or not on the table, and 
when something is a wine glass versus another kind 
of glass. But if every individual learns categories independently
and autonomously, how can they ever become shared? It is 
assumed that language helps in establishing a shared ontology within 
a language community but language itself depends on 
shared categories, so how can the whole system ever 
get off the ground; how can this chicken and egg situation
be broken?

I will begin this chapter by outlining the empiricist 
and rationalist positions, which have 
dominated much of the philosophical discussion on 
categorisation and have indirectly influenced attempts 
to build artificial systems able to perform some form
of categorisation. 
I will then propose an alternative {\itshape selectionist} approach
and study a categorisation system based on it. It is 
shown that agents endowed with this system are able
to develop an adequate repertoire of categories for 
distinguishing objects in their environment and that 
this repertoire remains adaptive when important changes 
occur in the environment. This goes some way to 
resolve the paradoxes of meaning, but the story remains
incomplete. To explain how agents can share the same 
categories even if they develop their ontologies 
independently of each other, I will later argue for
a co-evolution of language and meaning. I will show
that when ontological development is coupled to lexical 
development, the two become co-ordinated with neither
a central co-ordinator nor prior design.

@@

Rationalists claim that categories 
exist {\itshape a priori} and therefore categorisation comes from
within. They argue that humans have a repertoire of ideal universal
forms, which they project
onto reality. Reality itself is a weak, imperfect 
reflection of these forms, like the shadows of objects
on the wall of a cave. Because of this poverty of stimuli, 
categories (particularly the perceptually grounded
categories that are the focus of our attention in this 
book) are claimed to be unlearnable by induction and must
therefore be innate.\footnote{
Strong forms of innateness have been defended by 
\cite{Fodor:1983}
and many philosophers and linguists associated with the 
generative grammar paradigm. See the discussion in
\cite{Wierzbicka:1992}, particularly the introductory chapter. 
In artificial intelligence 
research, particularly the logic-oriented tradition, there
is often an implicit acceptance that basic categories 
are innate, but that derived categories, formulatable
in terms of more primitive concepts, can be learned
(see \citealt{McCarthy:2008}). There are also 
large-scale efforts going on to build an ontology as 
rich as human ontologies, see the discussions around
Doug Lenat's CYC project in: \cite{Steels:1994}.}

\is{innate categories}This innateness hypothesis suggests that the brain comes
with {\itshape categorisation organs}, small neuronal circuits
capable of performing 
the mapping of some idealised universal form onto
reality. Consequently the human genome
must include a set of {\itshape concept genes} which regulate how each of 
these categorisation organs should grow during 
development. Rationalists claim
that it is absurd to think of categories as being 
learned from example cases supplied by the environment, just 
as it is absurd to say that the hand learns to 
grow five fingers.

@@

There is something to say both for a rationalist and 
for an empiricist approach, indeed otherwise so many
serious thinkers could not have believed fervently 
in one position or the other. Rationalists point to the 
fact that children acquire concepts
surprisingly quickly and apparently with very little stimuli
and that anthropological observations have 
shown that there are strong universal tendencies
for basic perceptually grounded categories, such as colour or 
space. However, more detailed observations show
that the acquisition of categories in children goes 
in fact very slowly. For example, concepts like cause-effect
and the correct use of the word \smplenquote{because}, the proper
use of tenses, etc. all take years to develop. 
Adults keep acquiring
new categories through out their lifetime, which makes 
it difficult to maintain that they are part of the 
human genome. For example, 
airline pilots and sailors categorise the direction and 
strength of winds and the shapes and colours of clouds, 
so as to predict turbulences, future weather, or 
advantageous trajectories. There are
occasionally profound differences in how cultures conceptualise
reality. These differences do not seem to be innate because
everybody who has had sufficient exposure to the 
culture, preferably at an early age and therefore without too much 
preconception, can normally acquire them.\footnote{
Rationalists often argue that there is no other way 
to explain the child's rapid acquisition of concepts
but detailed psycholinguistic observations have shown that 
child language understanding (which is the 
clearest sign whether certain concepts have been acquired)
is often deceptive because 
non-verbal strategies may lead to appropriate answers
to adult questions and thus the appearance of 
understanding.}

@@

In a selectionist approach, a repertoire of primitive operations
is given, presumably implemented by the 
basic biochemistry of the neural systems, and there are
ways to combine these operations into more complex
{\itshape visual programs}. Those
programs yielding an outcome which
is afterwards used successfully in categorisation are 
retained and the others discarded, thus establishing 
a co-evolution between a repertoire of sensory channels
and a repertoire of 
categories. We have already done some successful experiments 
in this direction, see \cite{Dejong:1999},  
but this theme will not be pursued further here as a full discussion 
would be too much a digression from the main line
of investigation. For the remainder of this book,
the sensory channels will be pre-programmed, 
although it is still up to the selectionist 
categorisation process to discover which ones are useful
in the environments presented to the agents and which 
ones are not.

@@

Consider the category [\textsc{large}]. What do all large 
objects have in common? At first sight very little. 
Almost any physical object can be called \smplenquote{large} in 
one context or another. 
It is going to be very difficult for a learning agent
to determine some commonality, even if given thousands 
of examples of large objects. In the beginning, the agent
might be confused by commonalities in 
colour or position or shape. 
Only with some luck, will the learning algorithm
start to zoom in on size. 
This is what makes inductive learning so slow and 
why it is criticised, rightfully, by rationalists. 
In fact, searching for commonality hardly makes sense for
many categories. [\textsc{large}] is only meaningful in
opposition to [\textsc{small}], [\textsc{left}] only makes sense
in opposition to [\textsc{right}]. Often categorisation
is relative to the context (a block which is large when surrounded
by smaller blocks may be categorised as
small when surrounded by larger
blocks) and occasionally it is relative to the objects themselves
(a large mouse is always categorised as
much smaller than a small elephant).

@@

I will refer to the process capable of making a distinction as
a {\itshape categoriser}. %\cite{categoriser} #FIXME
It operates on the output of 
a sensory channel and decides whether a category or its opponent
is valid. A categoriser keeps track of its success 
by maintaining an internal counter. 
Consider for example the category [\textsc{large}] which 
operates on the output of the \textsc{area}-channel. This channel 
contains a scaled value 
between 0.0 and 1.0 for the area of a segment. A distinction 
can therefore be 
made simply by dividing the set of possible values in two
halves: those whose area is between
0.0 and 0.5 and those whose area is between 0.5 and
1.0, giving us two categories, [\textsc{small}] and [\textsc{large}]. When 
the agent needs to categorise an object, he checks in which
region an object's area falls. If it is between 0.0 and 0.5, the
category is [\textsc{small}], if it is greater than 0.5 the category
is [\textsc{large}].

@@

Still another way to categorise reality is by imposing 
an order on the segments based on their values for 
a particular sensory channel. For example, we can 
order the segments based on the \textsc{hpos} channel (i.e. 
from left to right) and then introduce relational categories
like \smplenquote{left of one segment in the series}, or \smplenquote{the left-most object}. 
Similarly we can order the segments based on the 
\textsc{height} channel (i.e. in terms of their size) and then 
have categories that select the smallest (i.e. the 
first segment in this ordering), or those greater than 
some other one.

>>>>

The previous chapter laid the perceptual groundwork 
for investigating conceptualisation and language 
communication. It proposed a set of mechanisms for segmenting a 
scene and extracting characteristics about each 
segment. The next task of the agent is to use these results
from the perceptual layer to categorise 
and hence conceptualise the
scene. For the time being, I will 
ignore the full complexity of natural
language meaning and focus only on the most 
simple type of conceptualisation one can
imagine, namely categories, which logicians refer 
to as unary predicates, as well as conjunctive 
combinations of categories. Words like \textit{blue}, \textit{light} or
\textit{square} name such categories. We not only need a way 
that the Talking Heads can use such categories to conceptualise
a scene based on visual perception, but we also need 
mechanisms that can explain how such categories might 
develop or be acquired by each agent autonomously and 
without supervised training. This is clearly an enormous
challenge and no known universally accepted solution 
exists to this problem.

@@

Then there is a third paradox, first suggested
by Jean-Jacques Rousseau. In order to communicate,
the speaker and the hearer need to share the building
blocks of the conceptualisations underlying their communication. 
If I say \textit{the wine glass on the table}, I expect 
the hearer to be able to recognise what objects are 
tables, when something is on the table or not on the table, and 
when something is a wine glass versus another kind 
of glass. But if every individual learns categories independently
and autonomously, how can they ever become shared? It is 
assumed that language helps in establishing a shared ontology within 
a language community but language itself depends on 
shared categories, so how can the whole system ever 
get off the ground; how can this chicken and egg situation
be broken?

I will begin this chapter by outlining the empiricist 
and rationalist positions, which have 
dominated much of the philosophical discussion on 
categorisation and have indirectly influenced attempts 
to build artificial systems able to perform some form
of categorisation. 
I will then propose an alternative {\scshape selectionist} approach
and study a categorisation system based on it. It is 
shown that agents endowed with this system are able
to develop an adequate repertoire of categories for 
distinguishing objects in their environment and that 
this repertoire remains adaptive when important changes 
occur in the environment. This goes some way to 
resolve the paradoxes of meaning, but the story remains
incomplete. To explain how agents can share the same 
categories even if they develop their ontologies 
independently of each other, I will later argue for
a co-evolution of language and meaning. I will show
that when ontological development is coupled to lexical 
development, the two become co-ordinated with neither
a central co-ordinator nor prior design.

@@

Rationalists claim that categories 
exist {\itshape a priori} and therefore categorisation comes from
within. They argue that humans have a repertoire of ideal universal
forms, which they project
onto reality. Reality itself is a weak, imperfect 
reflection of these forms, like the shadows of objects
on the wall of a cave. Because of this poverty of stimuli, 
categories (particularly the perceptually grounded
categories that are the focus of our attention in this 
book) are claimed to be unlearnable by induction and must
therefore be innate.\footnote{
Strong forms of innateness have been defended by 
\cite{Fodor:1983}
and many philosophers and linguists associated with the 
generative grammar paradigm. See the discussion in
\cite{Wierzbicka:1992}, particularly the introductory chapter. 
In artificial intelligence 
research, particularly the logic-oriented tradition, there
is often an implicit acceptance that basic categories 
are innate, but that derived categories, formulatable
in terms of more primitive concepts, can be learned
(see \citealt{McCarthy:2008}). There are also 
large-scale efforts going on to build an ontology as 
rich as human ontologies, see the discussions around
Doug Lenat's CYC project in \cite{Steels:1994}.}

\is{innate categories}This innateness hypothesis suggests that the brain comes
with {\scshape categorisation organs}, small neuronal circuits
capable of performing 
the mapping of some idealised universal form onto
reality. Consequently the human genome
must include a set of {\sc concept genes} which regulate how each of 
these categorisation organs should grow during 
development. Rationalists claim
that it is absurd to think of categories as being 
learned from example cases supplied by the environment, just 
as it is absurd to say that the hand learns to 
grow five fingers.

@@

There is something to say both for a rationalist and 
for an empiricist approach, indeed otherwise so many
serious thinkers could not have believed fervently 
in one position or the other. Rationalists point to the 
fact that children acquire concepts
surprisingly quickly and apparently with very little stimuli
and that anthropological observations have 
shown that there are strong universal tendencies
for basic perceptually grounded categories, such as colour or 
space. However, more detailed observations show
that the acquisition of categories in children goes 
in fact very slowly. For example, concepts like cause-effect
and the correct use of the word \textit{because}, the proper
use of tenses, etc. all take years to develop. 
Adults keep acquiring
new categories through out their lifetime, which makes 
it difficult to maintain that they are part of the 
human genome. For example, 
airline pilots and sailors categorise the direction and 
strength of winds and the shapes and colours of clouds, 
so as to predict turbulences, future weather, or 
advantageous trajectories. There are
occasionally profound differences in how cultures conceptualise
reality. These differences do not seem to be innate because
everybody who has had sufficient exposure to the 
culture, preferably at an early age and therefore without too much 
preconception, can normally acquire them.\footnote{
Rationalists often argue that there is no other way 
to explain the child's rapid acquisition of concepts
but detailed psycholinguistic observations have shown that 
child language understanding (which is the 
clearest sign whether certain concepts have been acquired)
is often deceptive because 
non-verbal strategies may lead to appropriate answers
to adult questions and thus the appearance of 
understanding.}

@@

In a selectionist approach, a repertoire of primitive operations
is given, presumably implemented by the 
basic biochemistry of the neural systems, and there are
ways to combine these operations into more complex
{\scshape visual programs}. Those
programs yielding an outcome which
is afterwards used successfully in categorisation are 
retained and the others discarded, thus establishing 
a co-evolution between a repertoire of sensory channels
and a repertoire of 
categories. We have already done some successful experiments 
in this direction, see \cite{Dejong:1999},  
but this theme will not be pursued further here as a full discussion 
would be too much a digression from the main line
of investigation. For the remainder of this book,
the sensory channels will be pre-programmed, 
although it is still up to the selectionist 
categorisation process to discover which ones are useful
in the environments presented to the agents and which 
ones are not.

@@

Consider the category [\textsc{large}]. What do all large 
objects have in common? At first sight very little. 
Almost any physical object can be called \textit{large} in 
one context or another. 
It is going to be very difficult for a learning agent
to determine some commonality, even if given thousands 
of examples of large objects. In the beginning, the agent
might be confused by commonalities in 
colour or position or shape. 
Only with some luck, will the learning algorithm
start to zoom in on size. 
This is what makes inductive learning so slow and 
why it is criticised, rightfully, by rationalists. 
In fact, searching for commonality hardly makes sense for
many categories. [\textsc{large}] is only meaningful in
opposition to [\textsc{small}], [\textsc{left}] only makes sense
in opposition to [\textsc{right}]. Often categorisation
is relative to the context (a block which is large when surrounded
by smaller blocks may be categorised as
small when surrounded by larger
blocks) and occasionally it is relative to the objects themselves
(a large mouse is always categorised as
much smaller than a small elephant).

@@

I will refer to the process capable of making a distinction as
a {\scshape categoriser}. %\cite{categoriser} #FIXME
It operates on the output of 
a sensory channel and decides whether a category or its opponent
is valid. A categoriser keeps track of its success 
by maintaining an internal counter. 
Consider for example the category [\textsc{large}] which 
operates on the output of the \textsc{area}-channel. This channel 
contains a scaled value 
between 0.0 and 1.0 for the area of a segment. A distinction 
can therefore be 
made simply by dividing the set of possible values in two
halves: those whose area is between
0.0 and 0.5 and those whose area is between 0.5 and
1.0, giving us two categories, [\textsc{small}] and [\textsc{large}]. When 
the agent needs to categorise an object, he checks in which
region an object's area falls. If it is between 0.0 and 0.5, the
category is [\textsc{small}], if it is greater than 0.5 the category
is [\textsc{large}].

@@

Still another way to categorise reality is by imposing 
an order on the segments based on their values for 
a particular sensory channel. For example, we can 
order the segments based on the \textsc{hpos} channel (i.e. 
from left to right) and then introduce relational categories
like \textit{left of one segment in the series}, or \textit{the left-most object}. 
Similarly we can order the segments based on the 
\textsc{height} channel (i.e. in terms of their size) and then 
have categories that select the smallest (i.e. the 
first segment in this ordering), or those greater than 
some other one.
